A thermodynamic view of dusty protoplanetary disks

### 2017-09-23 model.tex
tiny edits before second submission
updated to revision 1 submitted version

{\bf Here we consider Lagragian perturbations such that 
  \begin{align}
 P \to P + \real\left(\Delta P e^{-\ii\sigma t}\right)
 \end{align}
is the pressure of a fluid element (and similarly for $\rho$). 
 The Lagrangian perturbation $\Delta$ of a variable $X$ is $\Delta X = \delta X + \bm{\xi}\cdot\nabla X$
 and $\bm{\xi}$ is the Lagrangian displacement ($\xi_{x,z} =  \ii\dd
 v_{x,z}/\sigma$). Inserting the above pressure and density fields
 into Eq. \ref{work_def}, and noting that only products of
 perturbations contribute to $\mathcal{W}$ after time-averaging, we
 find for periodic oscillations (time dependence $e^{\ii\omega t}$ and
 real $\omega$) that:  
}
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

>>>>

{\bf Here we consider Lagragian perturbations such that 
  \begin{align}
 P \to P + \real\left(\Delta P e^{-\ii\sigma t}\right)
 \end{align}
is the pressure following a fluid element of the mixture (and similarly for $\rho$). 
 The Lagrangian perturbation $\Delta$ of a variable $X$ is $\Delta X = \delta X + \bm{\xi}\cdot\nabla X$
 and $\bm{\xi}$ is the Lagrangian displacement, so $\xi_{x,z} =  \ii\dd
 v_{x,z}/\sigma$. Inserting the above pressure and density fields
 into Eq. \ref{work_def}, and noting that only products of
 perturbations contribute to $\mathcal{W}$ after time-averaging, we
 find for periodic oscillations (time dependence $e^{\ii\omega t}$ and
 real $\omega$) that:  
}
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

### 2017-09-23 result.tex
updated to revision 1 submitted version

Fig. \ref{compare_vshear_varHd} shows that a radially-varying
dust-to-gas ratio increases the magnitude of the vertical shear rate
away from the midplane (see also Eq. \ref{vshear2}). Thus we {\bf
  typically} find higher VSI growth rates, as shown in
Fig. \ref{vsi_dust_varHd}. 
Surfaces modes, {\bf appearing at high $k_x$}, are more effectively
enhanced by the additional vertical shear induced by
$\p_r\epsilon$. {\bf We find that low frequency body modes with small
  $k_x$ are more affected by radial variations in the 
  dust-to-gas ratio.}

@@

{\bf We find in smooth disks large 
  $k_x$ is needed for appreciable growth rates, e.g. $k_xH_g=1800$
  with $\rhod/\rhog\propto r^{-1}$. 
  Local shearing box simulations \citep[e.g.][]{bai10b,yang16} may 
  be required to study the non-linear evolution of such short 
  wavelengths. Alternatively, as shown in Fig. \ref{vert_mixed_modes},
  a rapidly varying dust-to-gas ratio permits dynamical instability at
  longer radial wavelengths. 
}

>>>>

Fig. \ref{compare_vshear_varHd} shows that a radially-varying
dust-to-gas ratio increases the magnitude of the vertical shear rate
away from the midplane (see also Eq. \ref{vshear2}). Thus we {\bf
  typically} find higher VSI growth rates, as shown in
Fig. \ref{vsi_dust_varHd}. 
Surfaces modes, {\bf appearing at high $k_x$}, are more effectively
enhanced by the additional vertical shear induced by
$\p_r\epsilon$. {\bf Low frequency body modes with small
  $k_x$ are more affected by radial variations in the 
  dust-to-gas ratio than high-$k_x$, high-frequency body modes. }

@@

{\bf We find in smooth disks large 
  $k_x$ is needed for appreciable growth rates, e.g. $k_xH_g=1800$
  with $\rhod/\rhog\propto r^{-1}$. 
  Local shearing box simulations may 
  be required to study the non-linear evolution of such short 
  wavelengths \citep[e.g.][]{bai10b,yang16}. Alternatively, as shown in Fig. \ref{vert_mixed_modes},
  a rapidly varying dust-to-gas ratio permits dynamical instability at
  longer radial wavelengths, which might be resolvable in global disk 
  simulations. 
}

### 2017-09-23 summary.tex
updated to revision 1 submitted version

The main application we envision for the dusty/adiabatic gas analogy 
is to develop physical interpretations of dust-gas instabilities; and to find 
dusty analogs of pure gas instabilities in protoplanetary disks, 
which is discussed in \S\ref{dust_analogs}. One can also exploit the similarity to adapt existing
hydrodynamic codes to simulate dusty protoplanetary disks (see \S\ref{dust_sims}).

@@

However, the validity of the terminal velocity approximation depends
not just on $\tstop$, but also on the value of the dust gas-ratio, and
possibly the problem itself. 
%You can decide if this is a relevant improvement, and whether to
%include. 
For example, Table \ref{si_compare} shows the dust-rich
`linA' mode and dust-poor `linB' mode of the streaming instability
have the same $\tstop$, but the former is accurately captured by the
simplified equations, while the latter is not.  This suggests that for
the streaming instability the simplified equations is better suited
for $\rhod/\rhog>1$. The simplified equations may also contain spurious
modes in certain limits (see \S\ref{spurious_epi}).

@@

Note that our thermodynamic interpretation of dust-gas drift does
\emph{not} in fact depend on the terminal velocity approximation.   
Once the gas equation of state is fixed and the true energy equation deleted, 
the dust continuity equation can be converted to a new effective energy equation. 
As an example, for strictly isothermal dusty gas we obtain   
\begin{align}
\frac{DP}{Dt} = - P\nabla\cdot\bm{v} +
c_s^2\nabla\cdot\left[\fdust\left(1-\fdust\right)\rho\left(\bm{v}_\mathrm{d}
    - \bm{v}_\mathrm{g}\right)\right]. \label{no_term_vel}  
\end{align}
This equation does not use
the terminal velocity approximation (which gives
Eq. \ref{eff_energy}); evaluating the right-hand-side generally 
requires solving an evolutionary equation for $\bm{v}_\mathrm{d} - 
\bm{v}_\mathrm{g}$ \citep{laibe14}. Nevertheless,
Eq. \ref{no_term_vel} shows that dust-gas relative drift can be
interpreted as a heat flux within the dust-gas mixture. Thus, even 
without the terminal velocity approximation we can 
regard the dusty-gas as a single ideal fluid subject to cooling.   
}

@@

{\bf However, the dusty/adiabatic gas equivalence identified in this paper
means that one need not implement a separate equation for the dust component in
locally isothermal/polytropic gas. The standard energy equation substitutes for the dust continuity 
equation. Thus pure gas dynamic codes can be used to simulate protoplanetary dusty disks. 
} %means that pure 
%gas dynamics codes can be applied to simulate locally 
%isothermal/polytropic dusty gas. One need not implement a separate 
%equation for the dust component: 
%for strictly isothermal or polytropic gas perfectly-coupled to dust:  
% the standard energy equation substitutes for the dust continuity
%equation. 
When {\bf the sound-speed} $c_s$ (or $K$) is not constant and/or the dust-gas coupling is imperfect, $\tstop\neq0$, one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15} or thermal 
conduction, which is also common in modern hydrodynamics codes.

We have taken advantage of the dusty/adiabatic gas equivalence to 
convert the popular
\textsc{Pluto}\footnote{\url{http://plutocode.ph.unito.it/}}      
hydrodynamics code \citep{mignone07} into a dusty gas dynamics code appropriate for
simulating protoplanetary disks coupled to small dust. {\bf In fact, our code is unaware it is modeling dusty gas.} 
We will apply 
this modified code ---d\textsc{Pluto}--- to study dusty disk-planet interaction (Lin et al., 
in preparation). 
Preliminary results {\bf show our approach} %indicate the simplified one-fluid 
reproduce features such as dusty rings associated with 
planet-induced gaps similar to that obtained from explicit two-fluid
simulations \citep[e.g.][]{dong17}.

>>>>

The main application we envision for the dusty/adiabatic gas analogy 
is to develop physical interpretations of dust-gas drag instabilities; and to find 
dusty analogs of pure gas instabilities in protoplanetary disks, 
which is discussed in \S\ref{dust_analogs}. One can also exploit the similarity to adapt existing
hydrodynamic codes to simulate dusty protoplanetary disks (see \S\ref{dust_sims}).

@@

However, the validity of the terminal velocity approximation may not
only depend on $\tstop$. The value of the dust gas-ratio and the
problem itself may also be important. 
%You can decide if this is a relevant improvement, and whether to
%include. 
For example, Table \ref{si_compare} shows the dust-rich
`linA' mode and dust-poor `linB' mode of the streaming instability
have the same $\tstop$, but the former is accurately captured by the
simplified equations, while the latter is not.  This suggests that for
the streaming instability the simplified equations is better suited
for $\rhod/\rhog>1$. The simplified equations also contain spurious
modes in certain limits (see \S\ref{spurious_epi}).

@@

Note that our thermodynamic interpretation of dust-gas drift does
\emph{not} actually depend on the terminal velocity approximation.   
Once the gas equation of state is fixed and the true energy equation deleted, 
the dust continuity equation can be converted to a new effective energy equation. 
As an example, for strictly isothermal dusty gas we obtain   
\begin{align}
\frac{DP}{Dt} = - P\nabla\cdot\bm{v} +
c_s^2\nabla\cdot\left[\fdust\left(1-\fdust\right)\rho\left(\bm{v}_\mathrm{d}
    - \bm{v}_\mathrm{g}\right)\right]. \label{no_term_vel}  
\end{align}
This equation does not use
the terminal velocity approximation (which gives
Eq. \ref{eff_energy}). Evaluating the right-hand-side generally 
requires solving an evolutionary equation for $\bm{v}_\mathrm{d} - 
\bm{v}_\mathrm{g}$ \citep{laibe14}. Nevertheless,
Eq. \ref{no_term_vel} shows that dust-gas relative drift can be
interpreted as a heat flux within the dust-gas mixture. Thus, even 
without the terminal velocity approximation we can 
regard the dusty-gas as a single ideal fluid subject to cooling.   
}

@@

{\bf However, the dusty/adiabatic gas equivalence identified in this paper
means that one need not implement a separate module for simulating the dust component in
locally isothermal/polytropic gas. The standard energy equation substitutes for the dust continuity 
equation. Thus pure gas dynamic codes can be used to simulate protoplanetary dusty disks. 
} %means that pure 
%gas dynamics codes can be applied to simulate locally 
%isothermal/polytropic dusty gas. One need not implement a separate 
%equation for the dust component: 
%for strictly isothermal or polytropic gas perfectly-coupled to dust:  
% the standard energy equation substitutes for the dust continuity
%equation. 
When {\bf the sound-speed} $c_s$ (or $K$) is not constant and/or the dust-gas coupling is imperfect, $\tstop\neq0$, one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15} or thermal 
conduction, which is also common in modern hydrodynamics codes.

We have taken advantage of the dusty/adiabatic gas equivalence to 
convert the popular
\textsc{Pluto}\footnote{\url{http://plutocode.ph.unito.it/}}      
hydrodynamics code \citep{mignone07} into a dusty gas dynamics code appropriate for
simulating protoplanetary disks coupled to small dust. {\bf In fact,
  our code is unaware of the fact that it is modeling dusty gas.} 
We will apply 
this modified code ---d\textsc{Pluto}--- to study dusty disk-planet interaction (Lin et al., 
in preparation). 
Preliminary results {\bf show our approach} %indicate the simplified one-fluid 
reproduce features such as dusty rings associated with 
planet-induced gaps similar to that obtained from explicit two-fluid
simulations \citep[e.g.][]{dong17}.
